---
title: "Using Xkcd to Make Density Plots"
author: "Arthur Rocha"
date: "2019-04-24T12:03:06-03:00"
categories:
  - How to
tags:
  - R
  - xkcd
  - density
  - statistics
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = F,warning = F)
```

So, I've always been interesting in the ["xkcd type"](https://xkcd.com/) of figures only because **IT'S FUN**.


My goal here is to make density plots of the most common probability functions using the xkcd style.


- First of all, we will need to load these packages:
```{r}
library(xkcd)
library(dplyr)
library(ggplot2)
```

- Now, we do a function using the ggplot2 sintax to make the plot. Note that we need calculate the axes ranges and apply "jitter" to x and y to create a hand-drawn effect in the line. 

```{r}
xkcd_density = function(x,y){
  
  # Calcule the axis range
  xrange = range(x)
  yrange = range(y)
  
  # Plot line type + jitter
  qplot(jitter(x),jitter(y),geom = "blank")+
    geom_line(linetype = 1,size=1.5)+
    theme_xkcd()+
    xkcdaxis(xrange,yrange)+
    xlab("")+
    ylab("")
}
```


- Ploting the standart normal (0,1) density.

```{r}
x = seq(-3,3,by=.1)
y = dnorm(x,mean=0,sd=1)
xkcd_density(x,y)
```

- Ploting the Chi square (df = 3) density.

```{r}
x = c(0,seq(0.1,10,by=.1))
y = c(0,dchisq(x[-1],df=3))
xkcd_density(x,y)
```

There you have! A nice looking (sort of) xkcd density plot!